---
title: "Bootstrap inference in functional linear regression models with R"
author: 
  - name: Hyemin Yeon^[https://sites.google.com/view/hyeminyeon/]
date: '`r format(Sys.Date(), "%B %d, %Y")`'
bibliography: 
  - BTSinFLRMvign_bibfile.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrap inference in functional linear regression models with R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This is a brief introduction to the package <tt>`BTSinFLRM`</tt> based on Yeon, Dai, and Nordman [-@YDN23RB; -@YDN24PB; -@YDN25WB]. 
The package mainly contains three functions <tt>`RBinFLRM`</tt>, <tt>`PBinFLRM`</tt>, and <tt>`WBinFLRM`</tt>, 
which respectively implement the residual, paired, and (multiplier) wild bootstrap methods. 
The former is developed for functional linear regression models (FLRMs) under homoscedasticity while the latter two methods are used for heteroscedastic FLRMs.
We illustrate how to use the package with a simulated dataset.

## Generating Gaussian curves

We simulate independent pairs $\{(X_i, Y_i)\}_{i=1}^n$ of functional regressors and scalar responses based on the following FLRM:
$$Y = \langle \beta, X \rangle + \varepsilon.$$
The regressor curves $\{X_i\}_{i=1}^n$ are generated identically and independently from a Gaussian process. The construction is based on a truncated Karhuen--Loève expansion. The eigengap $\{\delta_j\}$ decays with polynomial decay rate $\delta_j = 2j^{-a}$ with $a=3.5$ and the eigenfunctions $\{\phi_j\}_{j=1}^J$ are chosen as Fourier functions. The slope function also has a polynomial decay rate $\beta_j = 2j^{-b}$ with $b=3.5$, and set $\beta = \sum_{j=1}^J \beta_j \phi_j$. Here, $J=15$ is used.
Then, the regressor is constructed as $X \overset{\mathsf{d}}{=} \sum_{j=1}^J \sqrt{\gamma_j} \xi_j \phi_j$ where $\{\gamma_j\}$ are eigenvalues and $\xi_j \sim \mathsf{N}(0,1)$ are (normalized) functional principal component scores. With uniform errors $\varepsilon_i \sim \mathsf{U}(0,1)$, the responses are generated as $Y_i = \langle \beta, X_i \rangle + \varepsilon_i$. 
```{r genData}
# simulate the regressor curves from Karhunen–Loève expansion and responses
J=15

# eigenvalues
c=2; a=3.5; dt = c*(1:J)^(-a)
g1 = c*VGAM::zeta(a)
ga = c(g1, sapply(1:(J-1), function(j){g1 - sum(dt[1:j])}))

# eigenfunctions
tt = 100; tGrid = seq(0, 1, len=tt); J=15
phi= t(fda::fourier(tGrid, J))

# slope function
b=3.5; bj = c*((1:J)^(-b))*(rbinom(J, 1, 0.5)*2-1)
beta = c(bj %*% phi)

# generate (Gaussian) regressors and responses
n=50; n0=3
xi = matrix(rnorm(n*J), n, J); X = xi %*% (phi * sqrt(ga))
xi0 = matrix(rnorm(n0*J), n0, J); X0 = xi0 %*% (phi * sqrt(ga))
er = runif(n); Y = c(X %*% beta / tt + er)
```


```{r, echo=FALSE, out.width='60%', fig.align='center', fig.width=7, fig.height=7}
matplot(tGrid, t(X), type='l', xlab="", ylab="",
        main="Generated Gaussian regressor curves")
```
















## Bootstrap inference

For illustration of bootstrap inference, we choose $k_n = 3$, $h_n \in \{4,5,6\}$, $g_n= \{4,5\}$ for involved truncation parameters and $M=100$ for bootstrap resample size.
The significance level is here $\alpha = 0.05$.

```{r}
# trunction parameters
kn_vec = c(3); hn_vec = c(4,5,6); gn_vec = c(4,5)

# residual bootstrap
M_bts=100

library(BTSinFLRM)
```


```{r, echo=FALSE, out.width='60%', fig.align='center', fig.width=7, fig.height=7}
resInfer = BTSinFLRM:::inferFLRM(X, Y, X0, tGrid, kn_vec, hn_vec, gn_vec)
tmx = max(c(kn_vec, hn_vec, gn_vec))
matplot(tGrid, t(resInfer$est$betaHat), type='l', xlab="", ylab="", 
        main="Estimated slope functions")
legend("bottomright", legend = paste0("kn=",1:tmx),
       lty=1:tmx, col=1:tmx, lwd=1)
```














### Residual bootstrap

This section describes inference by CLT and residual bootstrap under homoscedasticity [@YDN23RB].

```{r}
resRB = RBinFLRM(X, Y, X0, tGrid, kn_vec, hn_vec, gn_vec, M_bts)
```

<tt>`resRB$CLT`</tt> provides the confidence and prediction intervals based on central limit theorem [@YDN23RB] with a five dimensional array. Here, the truncation parameter $k_n$ involves estimation of the error variance $\sigma_\varepsilon^2$ with estimator $\hat{\sigma}_\varepsilon^2 \equiv n^{-1} \sum_{i=1}^n (Y_i - \langle \hat{\beta}_{k_n}, X_i \rangle)^2$. 

```{r}
resRB$CLT
```

<tt>`resRB$RB`</tt> provides different intervals based on residual bootstrap [@YDN23RB]. It is a list with three dimension where each dimension corresponds to different combinations of tuning parameters $(k_n, h_n, g_n)$. Each list contains a three dimensional array where the last dimension represent different types of intervals:

* Intervals for a truncated projection $\langle \Pi_{h_n} \beta, X_0 \rangle$, the projection $\langle \beta, X_0 \rangle$, and the predicted response $Y_0$. 
* Individual and simultaneous intervals.
    + Symmetrized and unsymmetrized individual intervals
    + Studentized and non-studentized simultaneous intervals.

```{r}
resRB$RB
# Access a component of list:
resRB$RB[["kn=3", "hn=5", "gn=4"]]
```















## Paired bootstrap

This section describes inference by CLT and paired bootstrap under heteroscedasticity [@YDN24PB].

```{r}
resPB = PBinFLRM(X, Y, X0, tGrid, kn_vec, hn_vec, gn_vec, M_bts)
```

<tt>`resPB$CLT`</tt> provides the confidence intervals based on central limit theorem [@YDN24PB] with a four dimensional array. Here, the truncation parameter $k_n$ is used to compute the estimated scaling $\hat{s}_{h_n}(X_{0,l})$ based on the observed samples.

```{r}
resPB$CLT
```

<tt>`resPB$PBintervals`</tt> provides different intervals based on paired bootstrap proposed by @YDN24PB. It is a list with three dimension where each dimension corresponds to different combinations of tuning parameters $(k_n, h_n, g_n)$ as <tt>`resRB$RB`</tt> provides above. Each list contains a four dimensional array where the last two dimensions represent different types of intervals:

* Intervals for a truncated projection $\langle \Pi_{h_n} \beta, X_0 \rangle$ and the untruncated projection $\langle \beta, X_0 \rangle$.
* Individual and simultaneous intervals denoted as <tt>`II`</tt> and <tt>`SI`</tt> respectively.
    + Symmetrized and unsymmetrized individual intervals approximated by either non-studentized or studentized bootstrap estimators.
    + Studentized and non-studentized simultaneous intervals.
    
For obtaining studentized simultaneous intervals, two different scaling terms can be constructed when we compute the bootstrap statistics. One is the estimated scaling $\hat{s}_{h_n}(X_{0,l})$ from the observed samples (denoted as <tt>`SI_std1`</tt>) while the other is computed as $\hat{s}^*_{h_n}(X_{0,l})$ from the bootstrap resamples (denoted as <tt>`SI_std2`</tt>). 

```{r}
resPB$PBintervals
# Access a component of list:
resPB$PBintervals[["kn=3", "hn=5", "gn=4"]]
```


<tt>`resPB$PBpvalues`</tt> contains the p-values from the bootstrap testing of whether the projection of the slope function $\beta$ onto the space spanned by the new regressors $\mathcal{X}_0 \equiv \{X_{0,l}\}_{l=1}^L$ based on the testing procedure described in Section 3 of @YDN24PB.
The null hypothesis $H_0$ is represented as $\Pi_{\mathcal{X}_0} \beta = 0$ where $\Pi_{\mathcal{X}_0}$ denotes the projection operator onto the space spanned by the new regressors $\mathcal{X}_0 \equiv \{X_{0,l}\}_{l=1}^L$.
This gives a list with three dimension where each dimension corresponds to different combinations of tuning parameters $(k_n, h_n, g_n)$.
Each dimension contains the p-values from different test statistics:

* The statistics either with or without studentization. The bootstrap statistics can be studentized by the estimated scaling terms from either samples or bootstrap resamples.
* $L^2$ or maximum type statistics.
* The test statistics are computed when either enforcing the null $H_0: \Pi_{\mathcal{X}_0} \beta = 0$ or not, which are respectively denoted as <tt>`tilde`</tt> and <tt>`hat`</tt>.

```{r}
resPB$PBpvalues
# Access a component of list:
resPB$PBpvalues[["kn=3", "hn=5", "gn=4"]]
```






## Wild bootstrap

This section describes inference by multiplier wild bootstrap under heteroscedasticity [@YDN25WB]. The results of <tt>`WBinFLRM`</tt> exhibit the same structure as the ones of <tt>`PBinFLRM`</tt>; the descriptions are hence omitted.

```{r}
resWB = WBinFLRM(X, Y, X0, tGrid, kn_vec, hn_vec, gn_vec, M_bts, multiplier="normal")
```

One notable difference is that users can choose multipliers used in <tt>`WBinFLRM`</tt> by sepcifying <tt>`multiplier`</tt> option.
Three types of multipliers can be considered: two-point distribution with $\mathsf{P}(W_i = -(\sqrt{5}-1)/2) = (\sqrt{5}+1) / (2\sqrt{5}) = 1-\mathsf{P}(W_i = (\sqrt{5}+1)/2)$, the standard normal distribution with $W_i \sim \mathsf{N}(0,1)$, and the distribution of $W_i = V_i/2+(V_i^2-1)/2$ for independent standard normal variables $\{V_i\}_{i=1}^n$, which gives the third moment one $\mathsf{E}^*[W_i^3]=1$.
These respectively correspond to <tt>`multiplier="bin"`</tt>, <tt>`multiplier="normal"`</tt>, and <tt>`multiplier="HM"`</tt>,
where the second type is chosen in this example.


```{r}
resWB$WBintervals
# Access a component of list:
resWB$WBintervals[["kn=3", "hn=5", "gn=4"]]
```

Although @YDN25WB does not contain the hypothesis tests based on the wild bootstrap,
one might mimic the procedure proposed by @YDN24PB in the wild bootstrap procedure.
The result <tt>`resWB$WBpvalues`</tt> below contains the p-values from this bootstrap testing procedure.


```{r}
resWB$WBpvalues
# Access a component of list:
resWB$WBpvalues[["kn=3", "hn=5", "gn=4"]]
```






## References
